---
title: ""
author: ""
date: ""
# Ensure links are colored and clickable
colorlinks: true
# Specify the color for URLs and other links (e.g., blue)
urlcolor: blue 
citecolor: blue
linkcolor: blue
output: 
  pdf_document:
    highlight: breezedark
    number_sections: true
    keep_tex: true
    latex_engine: xelatex
header-includes:
  - \usepackage[english]{babel}
  - \usepackage{natbib}
  - \usepackage{url}
  - \usepackage{amsmath}
  - \usepackage{graphicx}
  - \usepackage{parskip}
  - \usepackage{fancyhdr}
  - \usepackage{vmargin}
  - \setmarginsrb{3 cm}{2.5 cm}{3 cm}{2.5 cm}{1 cm}{1.5 cm}{1 cm}{1.5 cm}
  - \usepackage{float}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \rhead{FLXETH002 \& FRNBEN003}
  - \lhead{Analytics A2}
  - \cfoot{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2)
library(dplyr)
library(patchwork)
```

```{=tex}
\begin{titlepage}
    \centering
    % Ensure uct-logo.png is in your working directory or provide the full path
    \includegraphics[width = 6cm]{uct-logo.png}\\[1.0 cm]
    \vspace*{0.5 cm}
    \textsc{\LARGE University of Cape Town}\\[2.0 cm]
    \textsc{\Large STA4010W}\\[0.5 cm]
    \textsc{\large Analytics}\\[0.5 cm]
    \rule{\linewidth}{0.2 mm} \\[0.4 cm]
    { \huge \bfseries Assessment 2: Neural Networks}\\
    \rule{\linewidth}{0.2 mm} \\[1.5 cm]
    
    \begin{minipage}{0.4\textwidth}
        \begin{flushleft} \large
            \emph{Author:}\\
            Ethon Felix \\
            Benjamin Franco
            \end{flushleft}
            \end{minipage}~
            \begin{minipage}{0.4\textwidth}
            \begin{flushright} \large
            \emph{Student Number:} \\
            FLXETH002 \\
            FRNBEN003
        \end{flushright}
    \end{minipage}\\[2 cm]
    
    {\large 20 April 2026}\\[2 cm]
    \vfill
\end{titlepage}
```
\newpage

a.  Conduct an EDA by plotting input-response relationships and comment on whether a neural network is an appropriate model class.

\vspace{0.2cm}



```{r echo=TRUE, fig.width=4, fig.height=4, out.width="80%"}
# Read the data from the text file
data <- read.table("HittingTheWall_2026.txt", header = TRUE)


# Set up a 2 x 2 plotting grid
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# 1. Boxplot: Carbs vs Bombed
boxplot(Carbs ~ Bombed,
        data = data,
        col = c("blue", "red"),
        main = "Carbohydrate Intake by Bombing Status",
        xlab = "Bombed",
        ylab = "Carbohydrate intake (g/h)",
        border = "black")

# 2. Boxplot: Fluid vs Bombed
boxplot(Fluid ~ Bombed,
        data = data,
        col = c("#4E79A7", "#E15759"),
        main = "Fluid Intake by Bombing Status",
        xlab = "Bombed",
        ylab = "Fluid intake (L/h)",
        border = "black")

# 3. Boxplot: AveSpeed vs Bombed
boxplot(AveSpeed ~ Bombed,
        data = data,
        col = c("#4E79A7", "#E15759"),
        main = "Average Speed by Bombing Status",
        xlab = "Bombed",
        ylab = "Average speed (km/h)",
        border = "black")

# 4. Barplot: Proportion bombed by Sex
prop_bombed <- tapply(data$Bombed == "TRUE", data$Sex, mean)

barplot(prop_bombed,
        col = c("#76B7B2", "#F28E2B"),
        ylim = c(0, 1),
        main = "Proportion Bombed by Sex",
        xlab = "Sex",
        ylab = "Proportion bombed",
        border = "black")

# Add labels above bars
text(x = c(0.7, 1.9),
     y = prop_bombed,
     labels = round(prop_bombed, 2),
     pos = 3)

# Optional overall title
mtext("Exploratory Data Analysis of Bombing Risk", outer = TRUE, cex = 1.2, font = 2)
```

Exploratory data analysis reveals several important patterns in the dataset:

Carbohydrate Intake: The boxplot of Carbs intake versus Bombed status shows that runners who bombed out during the race tended to have lower carbohydrate intake per hour. There appears to be a threshold effect, where bombing is much more likely below a certain level of carb intake, supporting the need for a nonlinear model.

Fluid Intake: Similarly, the boxplot for Fluid intake indicates that bombing out is associated with lower fluid intake per hour. The relationship is not strictly linear, as there is considerable overlap and possible threshold effects.

Sex: The bar plot of Bombed proportion by Sex demonstrates that a higher proportion of males (67%) bombed compared to females (57%). This suggests that sex may play a role in the likelihood of bombing, and potential interactions should be considered.

Average Speed: The boxplot for AveSpeed shows little difference between bombed and non-bombed groups, indicating that average speed is not a strong predictor of bombing out in this sample.

The relationships between nutrition variables and bombing are not strictly linear, with overlapping distributions, threshold effects, and potential interactions. These findings suggest that a nonlinear model, such as a neural network, is appropriate for capturing the complexity of the data and providing more accurate predictions.
\newpage

b.  

To show that

$$
\sigma(z_j) = \underbrace{\frac{\exp(z_j)}{\sum_{i=1}^q \exp(z_i)}}_\textbf{Eq. 1} =\underbrace{\frac{\exp(z_j - z^*)}{\sum_{i=1}^q \exp(z_i - z^*)}}_{\textbf{Eq. 2}} \quad \text{where } z^* = \max \{z_1,z_2,\dots,z_q\}
$$ \vspace{0.2cm}

we start by expanding Equation 2 as follows

$$
\begin{aligned}
\frac{\exp(z_j - z^*)}{\sum_{i=1}^q \exp(z_i - z^*)} = \frac{\exp(z_j)\exp(-z^*)}{\sum_{i=1}^q \exp(z_i) \exp(- z^*)} = \frac{\exp(z_j)}{\sum_{i=1}^q \exp(z_i)}
\end{aligned}
$$

\vspace{0.2cm}

which shows that equations 1 and 2 are mathematically equivalent.

\vspace{0.2cm}

Subtracting the maximum logit $(z^*)$ ensures that the largest exponent is zero and all others are negative, preventing overflow in the exponential calculation. This makes the softmax calculation robust and reliable for all input values. This is especially important in neural networks and deep learning, where logits can be very large or very small.

c.  

```{r echo=TRUE}
softmax = function(X){
 max = apply(X, 2, max)
 X_centered = X
 for (i in 1:ncol(X)){
   X_centered[,i] = X[,i]-max[i]
 }
 
 exp_X = exp(X_centered)
 
 probs = exp_X
 col_sums = colSums(exp_X)
 for (i in 1:ncol(exp_X)){
   probs[,1] = exp_X[,i]/col_sums[i]
 }
 
 return(probs)
}
```

d.  

**Inputs:** - $X^u$: Demographic input matrix ($p^u \times N$) - $X^v$: Nutrition input matrix ($p^v \times N$)

**First Hidden Layer (Split):** - $W^u_1$: Weights for $u$-hemisphere ($m \times p^u$) - $b^u_1$: Biases for $u$-hemisphere ($m \times 1$) - $W^v_1$: Weights for $v$-hemisphere ($m \times p^v$) - $b^v_1$: Biases for $v$-hemisphere ($m \times 1$)

**Activations:** $$
A^u_1 = \tanh(W^u_1 X^u + b^u_1)
$$ $$
A^v_1 = \tanh(W^v_1 X^v + b^v_1)
$$ Both $A^u_1$ and $A^v_1$ are $m \times N$ matrices.

**Second Hidden Layer (Split):** - $W^u_2$: Weights for $u$-hemisphere ($1 \times m$) - $b^u_2$: Bias for $u$-hemisphere ($1 \times 1$) - $W^v_2$: Weights for $v$-hemisphere ($1 \times m$) - $b^v_2$: Bias for $v$-hemisphere ($1 \times 1$)

**Activations:** $$
A^u_2 = \tanh(W^u_2 A^u_1 + b^u_2)
$$ $$
A^v_2 = \tanh(W^v_2 A^v_1 + b^v_2)
$$ Both $A^u_2$ and $A^v_2$ are $1 \times N$ matrices.

**Output Layer:** - Combine $A^u_2$ and $A^v_2$ (e.g., stack to form a $2 \times N$ matrix) - $W_3$: Output weights ($q \times 2$) - $b_3$: Output bias ($q \times 1$)

$$Z^L = W_3 \begin{bmatrix} A^u_2 \\ A^v_2 \end{bmatrix} + b_3$$

$$Y = \text{softmax}(Z^L)$$

**Mapping to Standard Network:**

The split-brain network can be written as a standard $(2m, 2)$-network by concatenating the two input matrices and defining block-diagonal weight matrices for the first hidden layer, so that each block only receives its designated subset of inputs. The forward pass equations can be written in standard matrix notation, with clear shapes for each matrix.

**Summary Table of Dimensions:**

| Layer    | Matrix  | Dimensions     |
|----------|---------|----------------|
| Input    | $X^u$   | $p^u \times N$ |
| Input    | $X^v$   | $p^v \times N$ |
| Hidden 1 | $W^u_1$ | $m \times p^u$ |
| Hidden 1 | $W^v_1$ | $m \times p^v$ |
| Hidden 1 | $b^u_1$ | $m \times 1$   |
| Hidden 1 | $b^v_1$ | $m \times 1$   |
| Hidden 1 | $A^u_1$ | $m \times N$   |
| Hidden 1 | $A^v_1$ | $m \times N$   |
| Hidden 2 | $W^u_2$ | $1 \times m$   |
| Hidden 2 | $W^v_2$ | $1 \times m$   |
| Hidden 2 | $b^u_2$ | $1 \times 1$   |
| Hidden 2 | $b^v_2$ | $1 \times 1$   |
| Hidden 2 | $A^u_2$ | $1 \times N$   |
| Hidden 2 | $A^v_2$ | $1 \times N$   |
| Output   | $W_3$   | $q \times 2$   |
| Output   | $b_3$   | $q \times 1$   |
| Output   | $Z^L$   | $q \times N$   |
| Output   | $Y$     | $q \times N$   |

**Conclusion:**\
This structure shows how the split-brain network can be evaluated as a standard $(2m, 2)$-network, with appropriate expressions and matrix structures for each layer. The key is to keep the hemispheres separate in the first two layers, then combine their activation for the output layer, all within standard matrix-form updating equations.

e.  

```{r echo=TRUE}
X_u = cbind(data$AveSpeed, data$Sex)
X_v = cbind(data$Fluid, data$Carbs)

colnames(X_u) = c("AveSpeed", "Sex")
colnames(X_v) = c("Fluid", "Carbs")

Y <- cbind(
  as.numeric(data$Bombed),
  as.numeric(!data$Bombed)
)
```

f.  We count parameters layer by layer

\vspace{0.2cm}

**Weights**

\vspace{0.2cm}

-   Layer 1: $(p_u + p_v)m$

-   Layer 2: $2m$

-   Output: $2q$

Thus the total weight are

\vspace{0.2cm}

$$
(p_u + p_v+ 2)m + 2q
$$

**Biases:**

-   Layer 1: $2m$

-   Layer 2: $2$

-   Output: $q$

Thus total biases are

\vspace{0.2cm}

$$
2m + 2 + q
$$

Thus, the total parameters are

$$
(p_u + p_v + 4)m + 3q + 2
$$

For $m=4, p_u=p_v=2, q=2:$

$$
\text{Total parameters} = (2+2+4)\cdot4 + 3(2) + 2 = 32 + 6 + 2 = 40
$$

g.  

```{r}

```
